---
title: "Rmd Exercise Template"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Exercises Day 2
Richard Paquin Morel, adapted from exercises by Christina Maimone
2019-09-16

##Load the data
Load the gapminder dataset.

```{r}
gapminder <- read.csv(here::here("data", "gapminder5.csv"))
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
```


##If Statement
Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. 

Hint: use the any function.

```{r}
year <- 2002
if (any(gapminder$year == year) ) {
    print(paste("Yes, there are records from", year))
} else {
    print(paste("No records from", year))
}
```

Now do the same for 2012.

```{r}
year <- 2012
if (any(gapminder$year == year) ) {
    print(paste("Yes, there are records from", year))
} else {
    print(paste("No records from", year))
}
```


##Loop and If Statements
Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r}
pop_mean <- mean(gapminder$pop)
country <- unique(gapminder$country)
for (i in country) {
    
    if (mean(gapminder$pop[gapminder$country == i]) < pop_mean) {
        mean_le <- mean(gapminder$lifeExp[gapminder$country == i])
        print(paste("The mean life expectancy in", i, "is", mean_le))
    } 
}
```


Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}
conts <- unique(gapminder$continent)
for (i in conts) {
    
    mean_le_cont <- mean(gapminder$lifeExp[gapminder$continent == i])

    if (mean_le_cont < 50) {
        print(paste("Mean life expectancy of", i, " is less than 50."))
    } else if (mean_le_cont >= 50 & mean_le_cont <= 70) {
        print(paste("Mean life expectancy of", i, " is between 50 and 70."))
    } else {
        print(paste("Mean life expectancy of", i, " is greater than 70."))
    }
}
```


##Exercise: Write Functions
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.

```{r}
getInfo <- function (df) {
    name <- names(df)
    for (i in name) {
        print(paste(i, mode(df[1,i])))
    }
        
    
}
getInfo(gapminder)
```


Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.

```{r}
vecStats <- function (vect, include_median = FALSE) {
    vec_mean <- mean(vect)
    print(paste("The mean is", vec_mean))
    vec_sd <- sd(vect)
    print(paste("The sd is", vec_sd))
    if (include_median == TRUE) {
        vec_median <- median(vect)
        print(paste("The median is", vec_median))
    }
    
}

le <- gapminder$lifeExp
vecStats(le, include_median = F)
vecStats(le, include_median = T)
```


##Analyzing the relationship
Use what you???ve learned so far to answer the following questions using the gapminder dataset. Be sure to include some visualizations!

What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r}
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```



Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r}
viz_lm <-
  function(df, dv, iv, conts) {
    dat <- df[df[["continent"]] == conts, ]
    
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = conts,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }

for (i in country) {
    gapminder$continent[gapminder$country == "United States"] <- "North America"
    gapminder$continent[gapminder$country == "Canada"] <- "North America"
    gapminder$continent[gapminder$country == "Mexico"] <- "North America"
    
    
    
}

conts <- unique(gapminder$continent)

for (i in conts) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```

